Free Theorems for Nested Types

### 2020-06-25 free-theorems-nesteds/paper5.tex
added paper5.tex

\begin{abstract}
This paper considers parametricity and its consequent free theorems
for nested data types.  Rather than representing nested types via
their Church encodings into a higher-kinded extension of System F, we
adopt a functional programming perspective and design a polymorphic
calculus that provides primitives for constructing a robust class of
nested types, including ADTs and all (truly) nested types appearing in
the literature, directly as fixpoints \`{a} la Pitts' PolyPCF.  At the
term level, our calculus supports the usual pattern matching, map
functions, and stylized recursion via folds for all such types. Using
a variation on Atkey's 2012 construction
%of a parametric model
for $F_\omega$,
%that explicitly constructs equality operators rather than merely
%postulating them,
we construct a parametric model for our calculus. However, to ensure
the existence of the fixpoints interpreting nested types, cocontinuity
conditions on the functors underlying these fixpoints must be threaded
throughout our model construction. This turns out to be both delicate
and challenging. Overall, this paper offers a blueprint for
deriving principled parametricity results 
%constructing parametric models
for practical programming languages,
such as Haskell and Agda, that directly construct nested types as
fixpoints rather than encoding them. It also shows how parametric
models can be used to derive useful free theorems for programs over
nested types. Like their counterparts for ADTs, the free theorems our
model derives provide, in particular, a principled alternative to
proving properties of programs over nested types using, e.g., the deep
induction rules for nested types recently developed by Johann and
Polonsky. This is illustrated with several examples. 
\end{abstract}

@@

We introduce our calculus in Section~\ref{sec:calculus}.  At the type
level, it is the level-2-truncation of the higher-kinded calculus
from~\cite{jp19}, augmented with a primitive type of natural
transformations. To construct nested types, it constructs type
expressions not just from standard type variables, but also from type
constructor variables of various arities, and includes an explicit
$\mu$-construct for type-level recursion with respect to these
variables. The class of nested types thus constructed is very robust
and includes all (truly) nested types known from the literature. In
Section~\ref{sec:type-interp} we give set and relational
interpretations for the types From Section~\ref{sec:calculus}. As is
usual when modeling parametricity, types are interpreted as functors
from environments interpreting their type variable contexts to set or
relations, as appropriate. But in order to ensure that these functors
satisfy the cocontinuity properties needed to know that the fixpoints
interpreting $\mu$-types exist, set environments must map each $k$-ary
type constructor variable to an appropriately cocontinuous $k$-ary
functor on sets and relation environments must map each $k$-ary type
constructor variable to an appropriately cocontinuous $k$-ary relation
transformer, and these cocontinuity conditions must be threaded
throughout the type interpretations in such a way that the resulting
model is guaranteed to satisfy an appropriate Identity Extension Lemma
(Theorem~\ref{thm:iel}). Properly progagating the cocontinuity
conditions turns out to be both subtle and challenging, and
Section~\ref{sec:iel}, where it is done, is where the bulk of the work
in constructing our model lies.  At the term level, our calculus
includes primitive constructs for the actions on morphisms of the
functors interpreting types, initial algebras for fixpoints of these
functors, and structured recursion over elements of these initial
algebras (i.e., $\map$, $\tin$, and $\fold$ constructs,
respectively). While our calculus does not support general recursion
at the term level, it is strongly normalizing, so does perhaps edge us
toward the kind of provably total practical programming language
proposed at the end of~\cite{wad89}. In Section~\ref{sec:term-interp},
we give set and relational interpretations for the terms of our
calculus. As usual in parametric models, terms are interpreted as
natural transformations from interpretations of the term contexts in
which they are formed to the interpretations of their types, and these
must cohere in what is essentially a fibrational way. Immediately from
the definitions of our interpretations we prove in
Section~\ref{sec:Nat-type-terms} a scheme deriving free theorems that
are consequences of naturality of functions that are polymorphic over
nested types. This scheme is very general, is parameterized over both
the data type and the polymorphic function at hand, and has each of
the above $\mathtt{map}$-$\mathtt{reverse}$ theorems as instances. The
relationship between naturality and parametricity has long been of
interest, and our inclusion of a primitive type of natural
transformations allows us to clearly delineate those free theorems
that are consequences of naturality, and thus hold even in
non-parametric models, from those that require the full power of
parametricity. In Section~\ref{sec:thms} we prove that our model
satisfies an Abstraction Theorem (Theorem~\ref{thm:abstraction}), and
we derive several of this latter kind of free theorem from it in
Section~\ref{sec:ftnt}. Specifically, we state and prove
(non-)inhabitation results in Sections~\ref{sec:bottom}
and~\ref{sec:identity}, a free theorem for the type of a filter
function on generalized rose trees in Section~\ref{sec:ft-adt}, the
correctness of short cut fusion for lists in
Section~\ref{sec:short-cut}, and its generalization to nested types in
Section~\ref{sec:short-cut-nested}.

@@

Because our calclus is designed to ensure functoriality, and
functoriality does not interact with polymorphism as well as might be
hoped, our calculus cannot express nontrivial recursive functions, such
as a concatenation function for perfect trees, that take not just one
nested type parameterized over a variable $\alpha$ as input, but other
inputs parameterized over $\alpha$ as well. The fundamental issue is
that recursion is expressible only via folds, and it is well-known
that a fold over a nested type must take a natural transformation as
an argument and return a natural transformation as its result.

@@

can write when ADTs are viewed as proper nested types --- i.e., as
fixpoints of types interpreted as first-order functors --- and which
ones we can write when ADTs are viewed as nested types --- i.e., as
fixpoints of types that are interpreted as higher-order functors. This
is because enforcing functoriality requires that the arity of the
recursion variable $\phi$ in the fixpoint in the return type of
$\fold$ must equal the number of variables bound by its $\Nat$

@@

write as folds functions of types such as $\alpha; \emptyset \vdash
\Nat^\emptyset \ (\List\,\alpha)\,(\Nat^\emptyset \alpha\,(\List
    (\mu \beta. 1 + \beta)))$, all of whose argument types are
    parameterized over the same type $\alpha$. This is not possible
    for nested types, even when they are semantically
    equivalent to ADTs.

@@

We can, however, represent types of functions that take
multiple nested types as arguments, provided the types of those
arguments are parameterized over disjoint sets of type variables; the
return type of such a function can be parameterized over only 
the type variables occurring in its final argument's type.

@@

Moreover, even some recursive functions of a single proper

@@

with type signatures of the same problematic forms.

@@

Adding combinators for more sophisticated stylized recursion may
ultimately enhance the expressivity of our calculus but, as discussed
in Section~\ref{sec:conclusion}, the expressivity issue for folds for
nested types long pre-dates, and is somewhat orthogonal to, the issue
of parametricity for languages that construct nested types as
fixpoints studied here. Enhancing the expressivity of our calculus is
discussed further in Section~\ref{sec:conclusion}.
%, and is orthogonal to, our focus on
%constructing parametric models for calculi in which nested types are
%constructed as fixpoints.

The presence of the ``extra'' functorial variables in $\ol{\gamma}$ in
the rules for $\map^{\ol{F},\ol{G}}_H$, $\tin_H$, and $\fold^F_H$
merit special mention. They allow us to map or fold polymorphic
functions over nested types. Consider, for example, the polymorphic
function $\mathit{flatten} : \Nat^\beta
(\mathit{PTree}\,\beta)\,(\mathit{List}\,\beta)$ that uniformly maps
perfect trees to lists. Even in the absence of extra variables the
instance of $\map$ required to map each non-functorial monomorphic
instantiation of $\mathit{flatten}$ over a list of perfect trees is
well-typed:
\[\begin{array}{l}
\AXC{$\Gamma;\alpha \vdash \mathit{List} \, \alpha$}
\AXC{$\Gamma;\emptyset \vdash \mathit{PTree}\,F$}
\AXC{$\Gamma;\emptyset \vdash \mathit{List}\,G$}
\TIC{$\Gamma;\emptyset \,|\, \emptyset \vdash \map^{\mathit{PTree}\,F,
    \,\mathit{List}\,G}_{\mathit{List}\,\alpha} :
  \Nat^\emptyset (\Nat^\emptyset\,(\mathit{PTree}\, F)\,(\mathit{List}\, G))\,
  (\Nat^\emptyset\,(\mathit{List}\, (\mathit{PTree}\,
  F))\,(\mathit{List}\, (\mathit{List}\, G)))$} \DisplayProof
\end{array}\]
But in the absence of $\ol \gamma$, the instance
\[\Gamma;\emptyset~|~\emptyset \vdash
\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}
: \Nat^\emptyset ( \Nat^\beta(\mathit{PTree}\,
\beta)\,(\mathit{List}\, \beta))\, (\Nat^\beta\,(\mathit{List}\,
(\mathit{PTree}\, 
\beta))\,(\mathit{List}\, (\mathit{List}\, \beta)))\]
%\[\Gamma;\emptyset \,|\, x : \Nat^\beta(\mathit{PTree}\,
%\beta)\,(\mathit{List}\, \beta) \vdash
%\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}\,x
%: \Nat^\beta\,(\mathit{List}\,
%(\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
%\beta))\] of $\mathtt{map}$
required to map the {\em polymorphic}
$\mathit{flatten}$ function over a list of perfect trees is not: in
that setting the functorial contexts for $F$ and $G$ in the rule for
$\map^{F,G}_H$ would have to be empty, but the fact that the
polymorphic $\mathit{flatten}$ function is functorial in some
variable, say $\delta$, means that it cannot possibly have a type of
the form $\Nat^\emptyset F\, G$ that would be required for it to be
the function input to $\map$. Since untypeability of this instance of
$\map$ is unsatisfactory in a polymorphic calculus, where we naturally
expect to be able to manipulate entire polymorphic functions rather
than just their monomorphic instances, we use the ``extra'' variables
in $\ol \gamma$ to remedy the situation. Specifically, the rules from
Definition~\ref{def:well-formed-terms} ensure that the instance of
$\map$ needed to map the polymorphic $\mathit{flatten}$ function is
typeable as follows:
\[\begin{array}{l}
\AXC{$\Gamma;\alpha,\beta \vdash \mathit{List} \, \alpha$}
\AXC{$\Gamma;\beta \vdash \mathit{PTree}\,\beta \hspace*{0.3in}
  \Gamma;\beta \vdash \mathit{List}\,\beta$}
\BIC{$\Gamma;\emptyset~|~\emptyset \vdash
  \map^{F,G}_{\mathit{List}\,\alpha}  : \Nat^\emptyset
  (\Nat^\beta(\mathit{PTree}\, 
  \beta)\,(\mathit{List}\, \beta))\,
 (\Nat^\beta\,(\mathit{List}\,
  (\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
  \beta)))$}
\DisplayProof
\end{array}\]
Similar remarks explain the appearance of $\ol \gamma$ in the typing
rules for $\tin$ and $\fold$.

@@

The key to proving the Identity Extension Lemma
(Theorem~\ref{thm:iel}) in our setting is a familiar ``cutting down''
of the interpretations of universally quantified types, such as our
$\Nat$-types, to include only the ``parametric'' elements.
%(See, e.g.,~\cite{atk12,bfss90,gjfor15,rey83}).
This requires that set interpretations of types are defined
simultaneously with their relational interpretations. We give set
interpretations for our types in Section~\ref{sec:set-interp} and give
their relational interpretations in Section~\ref{sec:rel-interp}.
While the set interpretations are relatively straightforward, their
relation interpretations are less so, mainly because of the
cocontinuity conditions we must impose to ensure that they are
well-defined. We take some effort to develop conditions in
Section~\ref{sec:rel-interp}, which separates
Definitions~\ref{def:set-sem} and~\ref{def:rel-sem} in space, but
otherwise has no impact on the fact that they are given by mutual
induction.

@@

%A poset $\mcD = (D,\leq)$ is \emph{$\omega$-directed} if every finite
%subset of $D$ has an upper bound. When $\mcD$ is considered as a
%category, we write $d \in \mcD$ to indicate that $d$ is an object of
%$\mcD$ (i.e., $d \in D$). An {\em $\omega$-directed colimit} in a
%category $\mcC$ is a colimit of a diagram $F : {\mathcal D} \to \mcC$,
%where $\mathcal D$ is an $\omega$-directed poset. A category $\mcC$ is
%{\em $\omega$-cocomplete} if it has all $\omega$-directed colimits. A
%cocomplete category is one that has all colimits.
%
%If $\mcA$ and $\mcC$ are $\omega$-cocomplete, then the functor $F :
%\mcA \to \mcC$ is {\em $\omega$-cocontinuous} if it preserves
%$\omega$-directed colimits.  If $\mcA$ is locally small, then an
%object $A$ of $\mcA$ is {\em finitely presentable} if the functor
%$\mathsf{Hom}_\mcA(A, -) : \mcA \to \Set$ preserves $\omega$-directed
%colimits, i.e., if for every $\omega$-directed poset $\mathcal D$ and
%every functor $F : {\mathcal D} \to \mcC$, there is a canonical
%isomorphism $\colim{d \in \mcD}{\mathsf{Hom}_\mcA(A,Fd)} \simeq
%\mathsf{Hom}_{\mcA}(A, \colim{d \in \mcD}{Fd})$. A category $\mcA$ is
%       {\em finitely accessible} if it is $\omega$-cocomplete and has
%       a set $\mcA_0$ of finitely presentable objects such that every
%%       object is an $omega$-directed colimit of objects in $\mcA_0$;
%       it is {\em locally finitely presentable} if it is finitely
%       accessible and cocomplete.
We will interpret the types in our calculus as $\omega$-cocontinuous
functors on locally finitely presentable categories~\cite{ar94}. Since
functor categories of locally finitely presentable categories are
again locally finitely presentable, this will ensure, in particular,
that the fixpoints interpreting $\mu$-types in $\set$ and $\rel$
exist, and thus that both the set and relational interpretations of
all of the types in Definition~\ref{def:wftypes} are
well-defined~\cite{jp19}. To bootstrap this process, we interpret type
variables themselves as $\omega$-cocontinuous functors in
Definitions~\ref{def:set-env} and~\ref{def:reln-env}. If $\C$ and $\D$
are locally finitely presentable categories, we write $[\C,\D]$ for
the set of $\omega$-cocontinuous functors from $\C$ to $\D$.
%Note that the categories $\set$ and $\rel$ are both locally finitely
%presentable.

@@

\vspace*{-0.07in}

As in~\cite{jg10}, replacing $\onet$ with any type $\emptyset;\alpha
\vdash C$ generalizes
%with $\setsem{\emptyset;\alpha \vdash c}\rho = C$
Theorem~\ref{thm:short-cut-nested} to deliver more general a free
theorem whose conclusion is $\mathit{fold}_{H}\, B \; \circ \; G\; \mu
H \; \mathit{in}_{H} = G \,\setsem{\emptyset;\alpha \vdash K}\, B$.

\vspace*{0.02in}

It is also standard to prove that the parametric model constructed
verifies the existence of initial algebras. This is, however,
unnecessary here since initial algebras are built directly into our
model.

@@

In this paper we have constructed a parametric model for a calculus
providing primitives for constructing nested types directly as
fixpoints rather than representing them via their Chuch encodings. We
have also used the Abstraction Theorem for this model to prove free
theorems in the presence of nested types. This was not possible
before~\cite{jp19} because the nested types definable in our syntax
were not previously known to have well-defined interpretations in
locally finitely presentable categories like $\set$ and $\rel$. No
calculus for terms of these types existed, either. The key to
obtaining our parametric model is the delicate threading of
functoriality and its accompanying naturality conditions throughout
the model construction.

Note that all recursion in our calculus must be expressed using
folds. Expressivity of standard folds for nested types has long been
known to be a vexing issue; indeed, such folds can only express
computations that return natural transformations. Of course our
calculus inherits the constraints standard folds for nested types
impose, and thus is perhaps not as expressive as we would ideally
like. But since this expressivity issue is orthogonal to constructing
parametric models in the presence of nested types built from
primitives, and since it would detreact from the conceptual clarity of
our presentation, we have deliberately chosen not to address it in
this paper. (However, we discuss some ideas for increasing
expressivity in the next paragraph.) Despite the limitations of
standard folds, our results provide a blueprint for deriving
principled parametricity results for practical programming languages,
such as Haskell and Agda, that directly construct nested types as
fixpoints.

>>>>

\begin{abstract}
This paper considers parametricity and its consequent free theorems
for nested data types. Rather than representing nested types via their
Church encodings into a higher-kinded extension of System F, we adopt
a functional programming perspective and design a polymorphic calculus
that provides primitives for constructing a robust class of nested
types directly as fixpoints.
%\`{a} la Pitts' PolyPCF.
At the term level, our calculus supports
%the usual
pattern matching, map functions, and stylized recursion via folds for
all such types.
%Using a variation on Atkey's 2012 construction for $F_\omega$,
%
%of a parametric model
%that explicitly constructs equality operators rather than merely
%postulating them,
Our main technical contribution is the construction of a parametric
model for this calculus. This is both delicate and challenging.  To
establish a suitable Identity Extension Lemma for our calculus, mild
restrictions are imposed on the formation of fixpoints; all ADTs and
all (truly) nested types appearing in the literature can nevertheless
be expressed. Moreover, to ensure the existence of the fixpoints
interpreting nested types, cocontinuity conditions on the functors
underlying these fixpoints must be threaded throughout our model
construction.  Overall, this paper offers a blueprint for deriving
principled parametricity results for practical programming languages,
such as Haskell and Agda, that directly construct nested types as
fixpoints rather than encoding them. It also shows how parametric
models can derive useful free theorems for programs over nested
types. Like their counterparts for ADTs, these free theorems
%our model derives
provide
%, in particular,
a principled alternative to proving
properties of programs over nested types via induction, 
e.g., using the deep
induction rules
%for nested types
recently developed by
Johann and Polonsky.
%This is illustrated with several examples.
\end{abstract}

@@

We introduce our calculus in Section~\ref{sec:calculus}.  At the type
level, it is the level-2-truncation of the higher-kinded calculus
from~\cite{jp19}, augmented with a primitive type of natural
transformations.  To construct nested types, it carefully builds up
type expressions not just from standard type variables, but also from
type constructor variables of various arities, and includes an
explicit $\mu$-construct for type-level recursion with respect to
these variables. The class of nested types it constructs is very
robust and includes all (truly) nested types known from the
literature.  In Section~\ref{sec:type-interp} we give set and
relational interpretations
%{\color{blue} locally finitely presentable}
for the types from Section~\ref{sec:calculus}. As is usual when
modeling parametricity, types are interpreted as functors from
environments interpreting their type variable contexts to set or
relations, as appropriate. But in order to ensure that these functors
satisfy the cocontinuity properties needed to know that the fixpoints
interpreting $\mu$-types exist, set environments must map each $k$-ary
type constructor variable to an appropriately cocontinuous $k$-ary
functor on sets and relation environments must map each $k$-ary type
constructor variable to an appropriately cocontinuous $k$-ary relation
transformer, and these cocontinuity conditions must be threaded
throughout the type interpretations in such a way that the resulting
model is guaranteed to satisfy an appropriate Identity Extension Lemma
(Theorem~\ref{thm:iel}).
%{\color{blue} Syntactic restrictions on
%  $\mu$-types are essential to proving IEL.}
Properly progagating the cocontinuity conditions turns out to be both
subtle and challenging, and Section~\ref{sec:iel}, where it is done,
is where the bulk of the work in constructing our model lies.  At the
term level, our calculus includes primitive constructs for the actions
on morphisms of the functors interpreting types, initial algebras for
fixpoints of these functors, and structured recursion over elements of
these initial algebras (i.e., $\map$, $\tin$, and $\fold$ constructs,
respectively). While our calculus does not support general recursion
at the term level, it is strongly normalizing, so does perhaps edge us
toward the kind of provably total practical programming language
proposed at the end of~\cite{wad89}.  In
Section~\ref{sec:term-interp}, we give set and relational
interpretations for the terms of our calculus. As usual in parametric
models, terms are interpreted as natural transformations from
interpretations of the term contexts in which they are formed to the
interpretations of their types, and these must cohere in what is
essentially a fibrational way. Immediately from the definitions of our
interpretations we prove in Section~\ref{sec:Nat-type-terms} a scheme
deriving free theorems that are consequences of naturality of
functions that are polymorphic over nested types. This scheme is very
general, is parameterized over both the data type and the polymorphic
function at hand, and has each of the above
$\mathtt{map}$-$\mathtt{reverse}$ theorems as instances.  The
relationship between naturality and parametricity has long been of
interest, and our inclusion of a primitive type of natural
transformations allows us to clearly delineate those free theorems
that are consequences of naturality, and thus hold even in
non-parametric models, from those that require the full power of
parametricity.  In Section~\ref{sec:thms} we prove that our model
satisfies an Abstraction Theorem (Theorem~\ref{thm:abstraction}), and
we derive several of this latter kind of free theorem from it in
Section~\ref{sec:ftnt}.  Specifically, we state and prove
(non-)inhabitation results in Sections~\ref{sec:bottom}
and~\ref{sec:identity}, a free theorem for the type of a filter
function on generalized rose trees in Section~\ref{sec:ft-adt}, the
correctness of short cut fusion for lists in
Section~\ref{sec:short-cut}, and its generalization to nested types in
Section~\ref{sec:short-cut-nested}.

@@

Our rule for forming $\Nat$-types means that 
%Because our calculus is designed to ensure functoriality, and
%functoriality does not interact with polymorphism as well as might be
%hoped,
our calculus cannot express types of recursive functions, such as a
concatenation function for perfect trees, that take as inputs both a
nested type and an argument of another type, both of which are
parameterized over the same variable $\alpha$.
%not just one nested type parameterized over a variable $\alpha$ as
%input, but other inputs parameterized over $\alpha$ as well.
Similarly, our calculus cannot express the type of a zip function for
bushes. The fundamental issue is that recursion is expressible only
via folds, and a fold over a nested type must take a natural
transformation as its argument and return a natural transformation as
its result.

@@

can type when ADTs are viewed as ADTs --- i.e., as fixpoints of types
interpreted as first-order functors --- and which folds we can type
when ADTs are viewed as proper nested types --- i.e., as fixpoints of
types that are interpreted as higher-order functors --- because the
arity of the recursion variable $\phi$ in the fixpoint in the return
type of $\fold$ must equal the number of variables bound by its $\Nat$

@@

write types such as $\alpha; \emptyset \vdash \Nat^\emptyset
\ (\List\,\alpha)\,(\Nat^\emptyset \alpha\,(\List (\mu \beta. 1 +
\beta)))$, all of whose argument types are parameterized over the same
type $\alpha$. This is not possible even for nested types that are
semantically equivalent to ADTs.

@@

@@

Our calculus can, however, express types of functions
that take
multiple nested types as arguments, provided those nested types are
parameterized over disjoint sets of type variables and the return type of
the function is parameterized over only
the variables occurring in the type of its final argument.
Interestingly, even some recursive functions of a single proper

@@

with types of the same problematic forms.

@@

Allowing extra functorial variables in the codomains of $\Nat$-types
would recover typeability of, say, zip for bushes, but such extended
$\Nat$-types need not have semantics as $\omega$-cocontinuous functors
in $\set$, as will be required in Section~\ref{sec:type-interp}. For
example,
%as shown in Example D, Section VI of~\cite{jp19},
the type $\emptyset; \gamma \vdash \Nat^\emptyset (\onet + \onet)\,
\gamma$ does not. We would therefore have to construct our models
from semantic categories that are locally $\lambda$-presentable for
$\lambda > \omega$.
%It is tempting to try to recover typeability of, say, a zip function
%for bushes by allowing extra functorial variables in the codomain type
%of $\Nat$-types. But such extended $\Nat$-types need not have
%semantics as $\omega$-cocontinuous functors, as is required in
%Section~\ref{sec:type-interp}. For example, as shown in Example D,
%Section VI of~\cite{jp19}, the type $\emptyset; \gamma \vdash
%\Nat^\emptyset (\onet + \onet)\, \gamma$ does not.
Adding more expressive stylized recursion combinators could
also help,
%enhance the expressivity of our calculus
but the expressivity issue for folds for nested types long pre-dates,
and is somewhat orthogonal to, the issue of parametricity for
languages that construct nested types as fixpoints studied here.
Enhancing the expressivity of our calculus is discussed further in
Section~\ref{sec:conclusion}.

The presence of the ``extra'' functorial variables in $\ol{\gamma}$ in
the rules for $\map^{\ol{F},\ol{G}}_H$, $\tin_H$, and $\fold^F_H$
merit special mention. They allow us to map or fold polymorphic
functions over nested types. Consider, for example, the polymorphic
function $\mathit{flatten} : \Nat^\beta
(\mathit{PTree}\,\beta)\,(\mathit{List}\,\beta)$ that uniformly maps
perfect trees to lists. Even in the absence of extra variables the
instance of $\map$ required to map each non-functorial monomorphic
instantiation of $\mathit{flatten}$ over a list of perfect trees is
well-typed:
\[\begin{array}{l}
\AXC{$\Gamma;\alpha \vdash \mathit{List} \, \alpha$}
\AXC{$\Gamma;\emptyset \vdash \mathit{PTree}\,F$}
\AXC{$\Gamma;\emptyset \vdash \mathit{List}\,G$}
\TIC{$\Gamma;\emptyset \,|\, \emptyset \vdash \map^{\mathit{PTree}\,F,
    \,\mathit{List}\,G}_{\mathit{List}\,\alpha} :
  \Nat^\emptyset (\Nat^\emptyset\,(\mathit{PTree}\, F)\,(\mathit{List}\, G))\,
  (\Nat^\emptyset\,(\mathit{List}\, (\mathit{PTree}\,
  F))\,(\mathit{List}\, (\mathit{List}\, G)))$} \DisplayProof
\end{array}\]
But in the absence of $\ol \gamma$, the instance
\[\Gamma;\emptyset~|~\emptyset \vdash
\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}
: \Nat^\emptyset ( \Nat^\beta(\mathit{PTree}\,
\beta)\,(\mathit{List}\, \beta))\, (\Nat^\beta\,(\mathit{List}\,
(\mathit{PTree}\, 
\beta))\,(\mathit{List}\, (\mathit{List}\, \beta)))\]
%\[\Gamma;\emptyset \,|\, x : \Nat^\beta(\mathit{PTree}\,
%\beta)\,(\mathit{List}\, \beta) \vdash
%\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}\,x
%: \Nat^\beta\,(\mathit{List}\,
%(\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
%\beta))\] of $\mathtt{map}$
required to map the {\em polymorphic} $\mathit{flatten}$ function over
a list of perfect trees is not: in that setting the functorial
contexts for $F$ and $G$ in the rule for $\map^{F,G}_H$ would have to
be empty, but because the polymorphic $\mathit{flatten}$ function is
functorial in some variable, say $\delta$, it cannot possibly have a
type of the form $\Nat^\emptyset F\, G$ that would be required for it
to be the function input to $\map$. Since untypeability of this
instance of $\map$ is unsatisfactory in a polymorphic calculus, where
we naturally expect to be able to manipulate entire polymorphic
functions rather than just their monomorphic instances, we use the
``extra'' variables in $\ol \gamma$ to remedy the
situation. Specifically, Definition~\ref{def:well-formed-terms}
ensures that the instance of $\map$ needed to map the polymorphic
$\mathit{flatten}$ function is typeable as follows:
\[\begin{array}{l}
\AXC{$\Gamma;\alpha,\beta \vdash \mathit{List} \, \alpha$}
\AXC{$\Gamma;\beta \vdash \mathit{PTree}\,\beta \hspace*{0.3in}
  \Gamma;\beta \vdash \mathit{List}\,\beta$}
\BIC{$\Gamma;\emptyset~|~\emptyset \vdash
  \map^{F,G}_{\mathit{List}\,\alpha}  : \Nat^\emptyset
  (\Nat^\beta(\mathit{PTree}\, 
  \beta)\,(\mathit{List}\, \beta))\,
 (\Nat^\beta\,(\mathit{List}\,
  (\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
  \beta)))$}
\DisplayProof
\end{array}\]
Similar remarks explain the appearance of $\ol \gamma$ in the typing
rules for $\tin$ and $\fold$.

@@

The key to proving the Identity Extension Lemma
(Theorem~\ref{thm:iel}) in our setting is a familiar ``cutting down''
of the interpretations of universally quantified types, such as our
$\Nat$-types, to include only the ``parametric'' elements.
%(See, e.g.,~\cite{atk12,bfss90,gjfor15,rey83}).
This requires that set interpretations of types are defined
simultaneously with their relational interpretations. We give set
interpretations for our types in Section~\ref{sec:set-interp} and give
their relational interpretations in Section~\ref{sec:rel-interp}.
While the set interpretations are relatively straightforward, their
relation interpretations are less so, mainly because of the
cocontinuity conditions we must impose to ensure that they are
well-defined. We develop these conditions in
Section~\ref{sec:rel-interp}, which separates
Definitions~\ref{def:set-sem} and~\ref{def:rel-sem} in space, but
otherwise has no impact on the fact that they are given by mutual
induction.

@@

%A poset $\mcD = (D,\leq)$ is \emph{$\omega$-directed} if every finite
%subset of $D$ has an upper bound. When $\mcD$ is considered as a
%category, we write $d \in \mcD$ to indicate that $d$ is an object of
%$\mcD$ (i.e., $d \in D$). An {\em $\omega$-directed colimit} in a
%category $\mcC$ is a colimit of a diagram $F : {\mathcal D} \to \mcC$,
%where $\mathcal D$ is an $\omega$-directed poset. A category $\mcC$ is
%{\em $\omega$-cocomplete} if it has all $\omega$-directed colimits. A
%cocomplete category is one that has all colimits.
%
%If $\mcA$ and $\mcC$ are $\omega$-cocomplete, then the functor $F :
%\mcA \to \mcC$ is {\em $\omega$-cocontinuous} if it preserves
%$\omega$-directed colimits.  If $\mcA$ is locally small, then an
%object $A$ of $\mcA$ is {\em finitely presentable} if the functor
%$\mathsf{Hom}_\mcA(A, -) : \mcA \to \Set$ preserves $\omega$-directed
%colimits, i.e., if for every $\omega$-directed poset $\mathcal D$ and
%every functor $F : {\mathcal D} \to \mcC$, there is a canonical
%isomorphism $\colim{d \in \mcD}{\mathsf{Hom}_\mcA(A,Fd)} \simeq
%\mathsf{Hom}_{\mcA}(A, \colim{d \in \mcD}{Fd})$. A category $\mcA$ is
%       {\em finitely accessible} if it is $\omega$-cocomplete and has
%       a set $\mcA_0$ of finitely presentable objects such that every
%%       object is an $omega$-directed colimit of objects in $\mcA_0$;
%       it is {\em locally finitely presentable} if it is finitely
%       accessible and cocomplete.
We interpret types in our calculus as $\omega$-cocontinuous functors
on locally finitely presentable categories~\cite{ar94}. Since functor
categories of locally finitely presentable categories are again
locally finitely presentable, this ensures that the fixpoints
interpreting $\mu$-types in $\set$ and $\rel$ exist, and thus that
both the set and relational interpretations of all of the types in
Definition~\ref{def:wftypes} are well-defined~\cite{jp19}. To
bootstrap this process, we interpret type variables as
$\omega$-cocontinuous functors in Definitions~\ref{def:set-env}
and~\ref{def:reln-env}. If $\C$ and $\D$ are locally finitely
presentable categories, we write $[\C,\D]$ for the set of
$\omega$-cocontinuous functors from $\C$ to $\D$.
%Note that the categories $\set$ and $\rel$ are both locally finitely
%presentable.

@@

\vspace*{-0.08in}

As in~\cite{jg10}, replacing $\onet$ with any type $\emptyset;\alpha
\vdash C$ generalizes
%with $\setsem{\emptyset;\alpha \vdash c}\rho = C$
Theorem~\ref{thm:short-cut-nested} to deliver more general a free
theorem whose conclusion is $\mathit{fold}_{H}\, B \; \circ \; G\; \mu
H \; \mathit{in}_{H} = G \,\setsem{\emptyset;\alpha \vdash K}\, B$.
Although it is standard to prove that the parametric model constructed
verifies the existence of initial algebras, this is unnecessary here
since initial algebras are built directly into our model.

@@

We have constructed a parametric model for a calculus
providing primitives for constructing nested types directly as
fixpoints.
%rather than representing them via their Church encodings. We
We have also used the Abstraction Theorem for this model to prove free
theorems in the presence of nested types. This was not possible
before~\cite{jp19} because the nested types definable in our syntax
were not previously known to have well-defined interpretations in
locally finitely presentable categories like $\set$ and $\rel$. No
term calculus for terms of these types existed, either. The key to
obtaining our parametric model is
%{\color{blue} adjusting our type syntax to respect the accessibility
%  cardinal for the semantic category and}
the delicate threading of functoriality and its accompanying
naturality conditions throughout the model construction

Even when nested types have well-defined interpretations in a semantic
category, the types of standard functions over them may not be.  All
recursion in our calculus must be coded using folds.  Since standard
folds for nested types can only express computations that return
natural transformations, their expressivity has long been a vexing
issue, which our calculus naturally inherits.
%indeed, such folds
%Our calculus naturally inherits these constraints.
%Of course our calculus inherits the constraints
%standard folds for nested types impose, and thus is perhaps not as
%expressive as we would ideally like.
In addition, while there are many papers on higher-kinded types, we
know of none that considers expressiblity issues arising from the
local $\lambda$-presentability cardinal $\lambda$ of the categories in
which higher-kinded types are interpreted. Because $\lambda = \omega$
for $\set$ and $\rel$, the expressivity of our
%$\set$/$\rel$-based parametric model
calculus is not easily improved there. Construction of a
$\set$/$\rel$-based parametric model for our calculus is nevertheless
worthwhile.  For one thing, the expressivity issue is orthogonal to
constructing parametric models,
%in the presence of nested types built from primitives,
and trying to address both in a single paper would
detract from the conceptual clarity of our construction.
%our current focus on deriving principled parametricity
%results for practical programming languages, such as Haskell and Agda,
%that directly construct nested types as fixpoints.
%
%conceptual clarity of this paper.
For another, if the construction of this paper
%we couldn't
%construct a $\set$/$\rel$-based parametric model for the calculus
%presented here,
weren't possible then trying to construct a parametric model from
robust semantic categories for more data types and/or recursion
schemes would likely be futile. We naturally expect the construction
reported here to guide its generalization to the locally
$\lambda$-presentable categories for $\lambda > \omega$ that will be
needed to accommodate more general data types and more general
recursion over them.

% As is, the results reported here provide a
%blueprint for deriving principled parametricity results for practical
%programming languages, such as Haskell and Agda, that directly
%construct nested types as fixpoints.

%{\color{blue} Perhaps surprisingly, the accessibility cardinal
%$\omega$ for the semantic categories $\set$ and $\rel$ also impacts
%the types of functions (and thus the functions themselves), even over
%expressible data types, that can be expressed. Although there are
%many papers on higher-kinded types, no other work that we know of
%even considers expressiblity issues resulting from the semantic
%settings in which they are interpreted.}
%{\color{blue} In light of the aforementioned
%  restrictions imposed by the accessibility cardinal of $\set$ and
%  $\rel$, this paper is perhaps best read as a proof-of-concept for
%  constructing parametric models in the locally finitely presentable
%  categories $\set$ and $\rel$ for calculi that construct nested types
%  and other advanced data types as fixed points (rather than
%  representing them via their Church encodings) and that support
%  structured recursion over them. If we couldn't construct such a
%  model even in this restricted setting, then there'd be no point in
%  trying to do so for more robust categories, data types, and/or
%  recursion schemes. Moreover, our constructions in this simplest
%  possible setting will guide their generalizations to the locally
%  $\lambda$-presentable categories fopr $\lambda > \omega$ required to
%  handle more general data types, structured recursive functions over
%  then, and more general fixpoints.}
